A recipe for extremely reproducible enrichment analysis

### 2023-03-23 manuscript.Rmd
update ms

To address this, many articles have been written to endorse the use of "virtual machines" or
"Docker containers" to encapsulate the operating system and dependancies along with the research code
to facilitate a high degree of reproducibility [4-16](Perkel 2020; Stein 2010; Dudley & Butte 2010;
Sandve et al, 2013; Piccolo & Frampton 2016; Grüning et al, 2014; Lewis et al, 2016;
da Veiga Leprevost et al, 2017; Kulkarni et al, 2018; Pasquier et al, 2018; Hung et al, 2019;
Brito et al, 2020; Piccolo et al, 2021).
Despite the availability of such tools for over a decade, bioinformatics analyses using fully
reproducible virtual machines or containers remains a miniscule fraction.

In this article, we will focus on enrichment analysis, which is one of the most used methods in
computational biology, and involves the summarisation of omics data to reflect biological changes,
such as the detection of pathway activation in development or disease
[17,18](Slonim 2002; Khatri et al, 2012).
While enrichment analysis is undoubtedly a key method for the interpretation of large
data sets, there are concerns that inadvertant misuse of these methods leads to
unreliable results.
These potential problems include (i) the lack of correct background for over-representation tests,
(ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
[19-22](Timmons et al, 2005; Reimand et al, 2019; Wijesooriya et al, 2022; Zhao & Rhee 2023).
Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
the validity of the methods and results.
We are concerned about the reliability of published enrichment analyses, as web-based and
point-and-click tools appear to be far more popular than computer script based methods for this type
of work [21](Wijesooriya et al, 2022).
Web-based tools are a potential concern, algorithms and functional annotation sets undergo regular
updates, but the historical versions are typically not archived, nor are versions regularly reported
in publications, which hampers later reproducibility.
Indeed, in some cases the entire web resource can become unavailable, and hence irreproducible due to
a phenomenon known as link decay [23](Hennessey & Ge 2013).

@@

Therefore the most complete reproducibility solution is to use containerisation to provide a scripted
workflow with prescribed links to research data, external databases and complete record of parameter
selection.
Designed correctly, these containers could encapsulate bioinformatics workflows and ensure their
computational reproducibility for decades to come, with just one or a few simple commands.
There are multiple barriers to entry for such workflows including (i) the added difficulty compares to
existing point-and-click solutions; (ii) time poorness of researchers to learn new skills; and (iii)
there is a lack of step-by-step guides, and few existing templates which can be rapidly adapted.

@@

But is this extra work really necessary?
Although our previous work suggests that enrichment analyses on the whole lack reproducibility
[5](Wijesooriya et al, 2022), we lack the empirical data confirming this directly.
Therefore before recommending or prescribing Docker-based bioinformatics workflows, it is necessary
to assess directly whether current standards are appropriate, by trying to reproduce the enrichment
analyses of previously published studies.
In the following section we investigate the computational reproducibility of enrichment analyses
shown in a sample of 20 journal articles published in 2019.
This should provide a sound basis for recommendations.

>>>>

To address this, many articles have been written to endorse the use of "virtual machines" or
"Docker containers" to encapsulate the operating system and dependancies along with the research code
to facilitate a high degree of reproducibility [4-16](Perkel 2020; Stein 2010; Dudley & Butte 2010;
Sandve et al, 2013; Piccolo & Frampton 2016; Grüning et al, 2014; Lewis et al, 2016;
da Veiga Leprevost et al, 2017; Kulkarni et al, 2018; Pasquier et al, 2018; Hung et al, 2019;
Brito et al, 2020; Piccolo et al, 2021).
Despite the availability of such tools for over a decade, bioinformatics analyses using fully
reproducible virtual machines or containers remains only a small fraction.

In this article, we will focus on enrichment analysis, which is one of the most used methods in
computational biology, and involves the summarisation of omics data to reflect biological changes,
such as the detection of pathway activation in development or disease
[17,18](Slonim 2002; Khatri et al, 2012).
While enrichment analysis is undoubtedly a key method for the interpretation of large
data sets, there are concerns that inadvertant misuse of these methods leads to
unreliable results.
These potential problems include (i) the lack of correct background for over-representation tests,
(ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
[19-22](Timmons et al, 2005; Reimand et al, 2019; Wijesooriya et al, 2022; Zhao & Rhee 2023).
Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
the validity of the methods and results.
We are concerned about the reliability of published enrichment analyses, as web-based and
point-and-click tools appear to be far more popular than computer script based methods for this type
of work [21](Wijesooriya et al, 2022).
Web-based tools are a potential concern, firstly because sometimes methods are not completely described,
and secondly, algorithms and functional annotation sets undergo regular updates.
These historical versions are typically not archived, which hampers later reproducibility.
In some cases the entire web resource can become unavailable, and hence irreproducible due to
a phenomenon known as link decay [23](Hennessey & Ge 2013).

@@

Therefore the most complete reproducibility solution is to use containerisation to provide a scripted
workflow with prescribed links to research data, external databases and complete record of parameter
selection.
Designed correctly, these containers could encapsulate bioinformatics workflows and ensure their
computational reproducibility for decades to come, with just one or a few simple commands.
There are multiple barriers to entry for such workflows including (i) the added difficulty compared to
existing point-and-click solutions; (ii) time poorness of researchers to learn new skills; and (iii)
a lack of step-by-step guides and templates aimed at beginner users.

@@

But is this extra work really necessary?
Although our previous work suggests that enrichment analyses on the whole lack reproducibility
[5](Wijesooriya et al, 2022), we lack the empirical data confirming this directly.
Therefore before recommending or prescribing Docker-based bioinformatics workflows, it is necessary
to assess directly whether current standards are appropriate, by trying to reproduce the enrichment
analyses of previously published studies, and examining whether the conclusions of those articles
are supported or not.
In the following section we investigate the computational reproducibility of enrichment analyses
shown in a sample of 20 journal articles published in 2019.
This should provide a sound basis for recommendations.
